### 2025-04-26 code/Draft_script_2.qmd
d

---
title: "Data Science Applied to Ag - Final Project - ML"
format:
  html:
    embed-resources: true
    toc: true
    theme: cerulean
author: Md Shakir Moazzem, Umar Munir
---

# Start of data wrangling

# Introduction

This script contains the data wrangling steps for the final project.

# Setup

##Loading packages

The following code chunk will load necessary packages.

```{r Setup, message=F}

# Installing packages

#install.packages("tidyverse")
#install.packages("readxl")
#install.packages("janitor")
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("readr")
#install.packages("lubridata")
#install.packages("stringr")

# Loading packages

library(tidyverse)
library(readxl) # to read excel files
library(janitor) # to clean data; helps fix and standardize the column names
library(dplyr) # wrangling
library(tidyr) # wrangling
library(readr) # to export csv
library(lubridate)
library(stringr)

```

## Reading data

The following code chunk will read the csv files for the 3 training data sets

```{r training data import, message=F, warning=F}

#reading the csv files for the 3 training data sets

trait_df <- read_csv("../data/training/training_trait.csv") 
meta_df  <- read_csv("../data/training/training_meta.csv")
soil_df  <- read_csv("../data/training/training_soil.csv")

```

## EDA

```{r}

summary(trait_df)

View(trait_df)

```
```{r}

summary(meta_df)

```

```{r}

summary(soil_df)

```

# Data wrangling on "trait_df"

The code below creates a function to adjust the yield_mg_ha for 15.5% grain moisture.

```{r}

# Function to transform yield to 15.5% moisture
adjust_yield <- function(yield_mg_ha, grain_moisture) {
  yield_mg_ha * (100 - grain_moisture) / (100 - 15.5)
}

```

The code below conducts data wrangling on "trait_df"

```{r clean_and_summarize_trait, message=F, warning=F}

trait_clean <- trait_df %>%
  select(-block) %>%
  mutate(
    site = str_remove_all(site, "[a-z]"), # removing all lowercase letters from site
    site = str_replace(site, "-.*$", ""), # keeping only text before any dash
    site = str_replace(site, "_.*$", "") # keeping only text before any underscore
  ) %>%
  group_by(year, site, hybrid) %>%
  summarize(
    yield_mg_ha    = mean(yield_mg_ha, na.rm = TRUE),
    grain_moisture = mean(grain_moisture, na.rm = TRUE),
    date_planted   = first(date_planted),
    date_harvested = first(date_harvested),
    .groups = "drop"
  ) %>%
  mutate(
    adjusted_yield = adjust_yield(yield_mg_ha, grain_moisture),
    planting_date  = as.Date(date_planted,   "%m/%j/%y"),
    harvest_date   = as.Date(date_harvested, "%m/%j/%y"),
    plant_doy      = yday(planting_date),
    harvest_doy    = yday(harvest_date),
  ) %>%
  select(-yield_mg_ha, -grain_moisture, -date_planted, -date_harvested, -planting_date, -harvest_date) %>%
  ungroup()

trait_clean

```

```{r exporting wrangled data set}

#write_csv(trait_avg_df,
          #"../data/training_trait_clean.csv")
```

# Data wrangling on "soil_df"

The following code chunk will conduct data wrangling on  "training_soil.csv"

```{r data_wrangling_training_soil, message=F, warning=F}

soil_clean <- soil_df %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  ungroup()

soil_clean
```

```{r exporting wrangled data set}

#write_csv(soil_clean,
          #"../data/training_soil_clean.csv")
```

# Data wrangling on "meta_df"

The following code chunk will conduct data wrangling on  "training_meta.csv".

```{r clean_meta_sites}

meta_df2 <- meta_df %>%
  rename(
    lon = longitude,
    lat = latitude
  ) %>%
  mutate(
    # 1) remove all lowercase letters
    site = str_remove_all(site, "[a-z]"),
    # 2) keep only text before any dash
    site = str_replace(site, "-.*$", ""),
    # 3) keep only text before any underscore
    site = str_replace(site, "_.*$", "")
  ) %>%
  distinct(year, site, .keep_all = TRUE) %>%
  ungroup()

meta_df2

```

The code below will create categories for previous crops.

```{r}

meta_clean <- meta_df2 %>%
  mutate(
    # 1) lowercase & trim
    prev_crop_raw = str_to_lower(str_squish(previous_crop)),
    # 2) recode into your desired levels
    prev_crop = case_when(
      prev_crop_raw == "cotton"                                   ~ "cotton",
      prev_crop_raw == "peanut"                                   ~ "peanut",
      prev_crop_raw %in% c("soybean", "soybeans")                 ~ "soybean",
      prev_crop_raw == "sorghum"                                  ~ "sorghum",
      prev_crop_raw %in% c("wheat", "winter wheat", "2019/20 wheat") ~ "wheat",
      TRUE                                                         ~ "others"
    ) %>%
    # 3) make it a factor with the exact ordering you want
    factor(levels = c("cotton", "peanut", "soybean", "sorghum", "wheat", "others"))
  ) %>%
  select(-previous_crop, -prev_crop_raw) %>%
  clean_names()

meta_clean

```

# EDA for cleaned data

```{r}
summary(trait_clean)
```
```{r}
summary(soil_clean)
```

```{r}
summary(meta_clean)
```

```{r exporting wrangled data set}

#write_csv(meta_clean,
          #"../data/training_meta_clean.csv")
```

# Merging all 3 cleaned data frames

```{r}

merged_clean  <- trait_clean %>%
  left_join(soil_clean, by = c("year", "site")) %>%
  left_join(meta_clean, by = c("year", "site"))

merged_clean
```

```{r}
summary(merged_clean)
```

```{r exporting merged data set}

#write_csv(merged_clean,
          #"../data/training_merged_clean.csv")
```

# End of Data Wrangling

#Start of Open Source Daymet weather data download

The following code chunk will load necessary packages.

```{r Setup, message=F, warning=F}

# Installing packages

#install.packages("tidyverse")
#install.packages("sf") #to manipulate vector geospatial files
#install.packages("daymetr") #to retrieve data from daymet database website through R
#install.packages("remotes") #to install R packages that are not available on CRAN, and available on GitHub
#remotes::install_github("ropensci/USAboundaries") 
#remotes::install_github("ropensci/USAboundariesData")

# Loading packages

library(tidyverse) #need to load "tidyverse" package at first
library(sf) # for US map #to manipulate vector geo-spatial points
library(daymetr) #to retrieve data from daymet database website through R
library(remotes)
library(USAboundaries) # for US state boundaries
library(USAboundariesData)

```

The following code chunk will create a map of the USA and plot the sites in the map based on their latitude and longitude.

```{r create map of USA and add points, message=F, warning=F}

states <- us_states() %>% 
  filter( !(state_abbr %in% c("PR", "AK", "HI")) ) #to remove "PR" (Puerto Rico), "AK" (Alaska), and "HI" (Hawaii)

ggplot() +
  geom_sf(data = states) + #"geom_sf()" is used to plot "sf" object, which we just created above as "states" object; plots all states and territories of USA
  geom_point(data = merged_clean,
             aes(x = lon, #"Longitude" goes on longitude
                 y = lat) #"Latitude" goes on latitude
             ) +
  labs(
    title = "Corn Trial Site Locations (2014–2023)",
    x     = "Longitude",
    y     = "Latitude"
  )

```

The following code chunk will keep the observations for site-years having latitude and longitude withing the Daymet range of co-ordinates.

```{r}

# Define Daymet bounding box (WGS-84)
min_lat <- 14.53
max_lat <- 52.00
min_lon <- -131.104
max_lon <- -52.95

# Filter merged to Daymet’s valid range, dropping any NA coords
merged_daymet <- merged_clean %>%
  filter(
    !is.na(lat),
    !is.na(lon),
    lat  >= min_lat,
    lat  <= max_lat,
    lon  >= min_lon,
    lon  <= max_lon
  )

# Report how many rows remain (and were dropped)
message("Rows kept: ", nrow(merged_daymet), 
        " (dropped: ", nrow(merged_clean) - nrow(merged_daymet), ")")

```
The following code chunk will extract unique combinations of year, site, and their coordinates.

```{r unique_site_years_with_coords}

site_year_df <- merged_daymet %>%
  select(year, site, lon, lat) %>%  # need to include longitude and latitude along with site-years
  distinct() %>%                    
  arrange(year, site)

site_year_df

```

The following code chunk will download the weather data for all unique combinations of year, site, and their coordinates in the "site_year_df" object.

```{r}

weather_daymet_all <- site_year_df %>% 
  mutate(weather = pmap(list(.y = year, 
                             .site = site, 
                             .lat = lat, 
                             .lon = lon), 
                        function(.y, .site, .lat, .lon) 
                          download_daymet( 
                            site = .site, 
                            lat = .lat, #specifying ".lat" placeholder for "lat = " argument
                            lon = .lon, #specifying ".lon" placeholder for "lon = " argument
                            start = .y, 
                            end = .y, 
                            simplify = T,
                            silent = T) %>% #end of " download_daymet()" function
                          rename(.year = year,
                                 .site = site) 
                        ))

weather_daymet_all

```

The following code chunk will unnest the weather column.

```{r}

weather_daymet_unnest <- weather_daymet_all %>%
  select(year, site, weather) %>% 
  unnest(weather) %>% 
  pivot_wider(names_from = measurement, 
              values_from = value) %>% 
  janitor::clean_names()

weather_daymet_unnest

View(weather_daymet_unnest)

```

# Merging the weather data retrieved from Daymet to the "merged_clean" data

The following code chunk will merge the weather data retrieved from Daymet to the "merged_clean" data

```{r merge_weather_with_data}

# Join daily weather onto your cleaned trial data by year & site

daymet_all_unnest <- merged_clean %>%
  left_join(
    weather_daymet_unnest,
    by = c("year", "site")
  )

fieldweather <- daymet_all_unnest

fieldweather

View(fieldweather)

```

## Exporting the merged data

```{r}

#write_csv(daymet_all_unnest,
          #"../data/merged_fieldweatherdata.csv"
          #)

```

# End of retreiving weather data from Daymet

# Start of feature engineering

## Setup

```{r}

#install.packages("ggridges")

library(ggridges)
library(tidyverse)

```

```{r}

#fieldweather <- read_csv("../data/merged_fieldweatherdata.csv")

#fieldweather

```

The following code chunk will keep desired variables that we will use further and get abbreviated month name based on the date.

```{r}

fe_month <- fieldweather %>%
  # Selecting needed variables
  dplyr::select(year, 
                site, 
                hybrid, 
                lon, 
                lat, 
                yday, 
                yield = adjusted_yield, 
                prev_crop,
                plant_doy,
                harvest_doy,
                soil.ph = soilpH, 
                soil.om.pct = om_pct, 
                soil.k.ppm = soilk_ppm, 
                soil.p.ppm = soilp_ppm, 
                dayl.s = dayl_s, #to rename variable name from "dayl_s" to dayl.s
                prcp.mm = prcp_mm_day, #to rename variable name to "prcp.mm"
                srad.wm2 = srad_w_m_2,#to rename variable name to "srad.wm2"
                tmax.c = tmax_deg_c, #to rename variable name to "tmax.c"
                tmin.c = tmin_deg_c,#to rename variable name to "tmin.c"
                vp.pa = vp_pa #to rename variable name to "vp.pa"
                ) %>%
  # Creating a date class variable  
  mutate(date_chr = paste0(year, "/", yday)) %>% 
  mutate(date = as.Date(date_chr, "%Y/%j")) %>% 
  # Extracting month from date  
  mutate(month = month(date)) %>% 
  mutate(month_abb = month(date, label = T)) #To get abbreviated month name e.g., Jan, Feb, Mar,...,Dec #month_abb is in "ord" (ordinal) format

fe_month
```

The following code chunk will feature engineer planting and harvest stages based on the following cut-offs:

Planting
•	early: DOY ≤ 120
•	normal: 121–150
•	late: > 150

Harvest
•	early: DOY ≤ 275
•	normal: 276–305
•	late: > 305

```{r}

fe_month <- fe_month%>%
  mutate(
    planting_stages = case_when(
      plant_doy <= 120                     ~ "early",
      plant_doy >= 121 & plant_doy <= 150  ~ "normal",
      plant_doy > 150                      ~ "late"
    ) %>% 
      fct_relevel("early","normal","late"),

    harvest_stages = case_when(
      harvest_doy <= 275                    ~ "early",
      harvest_doy >= 276 & harvest_doy <= 305 ~ "normal",
      harvest_doy > 305                     ~ "late"
    ) %>% 
      fct_relevel("early","normal","late")
  ) %>%
  select(-plant_doy, -plant_doy)

fe_month

```

The following code chunk will summarize daily weather variables based on month.

```{r fe_month_sum}

fe_month_sum <- fe_month %>%
  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame 
  #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
  summarise(across(.cols = c(soil.ph, 
                             soil.om.pct, 
                             soil.k.ppm, 
                             soil.p.ppm,
                             dayl.s,
                             srad.wm2,
                             tmax.c,
                             tmin.c,
                             vp.pa),
                   .fns = mean, #do not indicate the actual function "mean()", just use the word "mean"
                   .names = "mean_{.col}"), #specifying the weather variables that we want their mean as new column variables #1st across() is applying the "mean" function (to summarize "mean")
            across(.cols = prcp.mm,
                   .fns = sum,
                   .names = "sum_{.col}"
                   ) #specifying the weather variable (prcp.mm) that we want its sum as new column variable #2nd across() is summarizing sum #2nd across() is applying the "sum" function (to summarize "sum")
            ) %>%
  ungroup() #To convert from "group" to "tibble"

fe_month_sum

```

The following code chunk will check "tmax.c" and "prcp.mm" for the first site-year and month for the purpose of double checking to make sure that we did everything okay the way we intended.

```{r}

fe_month %>%
  filter(year == 2014 & 
           site == "DEH1" &
           hybrid == "B37/MO17" &
           month_abb == "Jan") %>%
  summarise(tmax.c = mean(tmax.c),
            prcp.mm = sum(prcp.mm))

```

The code below will put the month as part of the column name.

```{r fe_month_sum_wide}

fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% 
  mutate(varname = paste0(name, "_", month_abb)) %>% 
  dplyr::select(-name, -month_abb) %>% 
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(10:ncol(.), ~round(., 1) )) %>%
  select(-lon, -lat, -mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)

fe_month_sum_wide

View(fe_month_sum_wide)

```

# EDA round 2  
Let's make a ridge plot to visualize the distribution of one variable over months.

```{r, message=FALSE, warning=FALSE}

#install.packages("ggridges")
library(ggridges) #really powerful package to plot distributions e.g., density plot

ggplot(data = fe_month_sum,
       aes(x = mean_tmax.c,
           y = month_abb,
           fill = stat(x) #fill = stat(x): specific to "ggridges"
           )
       ) +
  geom_density_ridges_gradient(scale = 3,
                               rel_min_height = 0.01) + #"the 2nd argument is"rel_min_height = 0.01": to cut down the tails of the distribution to look nice
  scale_fill_viridis_c(option = "C") + #"viridis"scale_fill_viridis_c()": color-blind-friendly color scale #argument (option = "C") to change the color to "magma" because magma is weather related; "option = " varies from A to F
  theme(legend.position = "none") #"theme(legend.position = "none")": to remove the legend from the graph

```

Now let's do that for all weather variables [ we will automate using map2() to conduct iteration ]

```{r}

finalplots <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% #we need to do pivot_longer() to iterate over the weather variables
  group_by(name) %>%
  nest() %>% #for iteration, we use "group_by()" followed by "nest()" in combo
  #we will use map2() since we need to iterate over 2 columns
  mutate(plot = map2(data, name, #map2() takes 2 arguments: the 1st argument becomes .x, the 2nd argument becomes .y #we must use map2() with a mutate() at first for iteration #"map2(data, name" : we want to iterate over "name" for "data"
                     ~ ggplot( data = .x, # .x represent "data" in the map2() function, so we need to use data = .x [= "data" from map2() ] to feed the "data" from map2() function as the data of ggplot() #.x is the iterating column of map2() that is a place holder for the 1st argument of map2()
       aes(x = value,
           y = month_abb,
           fill = stat(x)
           )
       ) +
  geom_density_ridges_gradient(scale = 3,
                               rel_min_height = 0.01) + 
  scale_fill_viridis_c(option = "C") + 
  theme(legend.position = "none") +
  labs(x = .y) #to rename the x-axis as the variable name # .y is the placeholder for "name" (i.e., variable name) in map2() function
                     ))

finalplots

```

```{r}

finalplots$plot #$plot to print all the ggplots for each variables

```

```{r}

#write_csv(fe_month_sum_wide,
          # "../data/weather_monthsum.csv")

```

# End of feature engineering

# Start of XGboost

# ML Workflow with XGBoost

## Step 1: Load Libraries

We begin by loading the required libraries:

```{r}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage
library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   # For parallel computing (useful during resampling/tuning)
#library(caret)       # Other great library for Machine Learning 
```

The following code chunk will rename the "fe_month_sum_wide" data frame as "weather".

```{r}

weather <- fe_month_sum_wide
weather

```

# ML workflow

## 1. Pre-processing

```{r weather_split}

set.seed(931735) # Setting seed to get reproducible results

weather_split <- initial_split(
  weather, 
  prop = .7, # proption of split same as previous codes
  strata = yield  # Stratify by target variable
  )

weather_split

```

### a. Data split

For data split, let's use **70% training / 30% testing**.

```{r weather_train}

weather_train <- training(weather_split)  # 70% of data

weather_train #This is your traing data frame

```

```{r weather_test}

weather_test <- testing(weather_split)    # 30% of data

weather_test

```

### b. Distribution of target variable

```{r distribution}
ggplot() +
  geom_density(data = weather_train, 
               aes(x = yield),
               color = "red") +
  geom_density(data = weather_test, 
               aes(x = yield),
               color = "blue")

```

### c. Data processing with recipe

Creating a recipe is an easy way to port your processing steps for other data sets without needing to repeat code, and also only considering the data it is being applied to.

You can find all available recipe step options here: https://tidymodels.github.io/recipes/reference/index.html

Different model types require different processing steps.\
Let's check what steps are required for an elastic net model (linear_reg). We can search for that in this link: https://www.tmwr.org/pre-proc-table #Don't need for numeric data

```{r weather_recipe}

# Create recipe for data preprocessing
weather_recipe <- recipe(yield ~ ., data = weather_train) %>% # Remove identifier columns and months not in growing season
  step_rm(
    year,       # Remove year identifier
    site,       # Remove site identifier
    matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
  ) %>%
  # Turning every factor (prev_crop, planting_stages, harvest_stages, etc.) into dummies
  step_dummy(all_nominal_predictors()) %>%
  # Imputation for any numeric NAs   
  step_impute_median(all_numeric_predictors())

weather_recipe

```

```{r weather_prep}
# Prep the recipe to estimate any required statistics
weather_prep <- weather_recipe %>% 
  prep()

# Examine preprocessing steps
weather_prep
```
## 2. Training

### a. Model specification

First, let's specify:\
- the **type of model** we want to train\
- which **engine** we want to use\
- which **mode** we want to use

> Elastic nets can only be run for a numerical response variable. XgBoost can be run with either numerical (regression) or categorical (classification) explanatory variable. Therefore, we have the need to specify the mode here.

XgBoost **hyperparameters**:

-   **`trees`**: The number of boosting rounds (i.e., how many trees will be added sequentially).\
-   **`tree_depth`**: Controls how deep each individual tree can grow. Deeper trees can capture more complex interactions but also increase the risk of overfitting.\
-   **`min_n`**: Minimum number of observations required in a node for it to be split. Acts as a regularization tool to prevent overly specific splits.\
-   **`learn_rate`**: Also known as `eta`, it controls how much each additional tree contributes to the overall model. Smaller values make the model more stable but require more trees.

```{r xgb_spec}

xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),  # Maximum depth of each tree
  min_n = tune(),  # Minimum samples required to split a node
  learn_rate = tune()) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
  set_engine("xgboost") %>% #specify engine 
  set_mode("regression")  # Set to mode
   # Total number of boosting iterations

        # Step size shrinkage for each boosting step

xgb_spec

```

### b. Cross-validation setup

We use 5-fold cross-validation to evaluate model performance during tuning:

```{r}

set.seed(235) #34549

resampling_foldcv <- vfold_cv(weather_train, 
                              v = 5)

resampling_foldcv

resampling_foldcv$splits[[1]]

```
### c. Hyperparameter grid with Latin Hypercube Sampling

We use Latin hypercube sampling to generate a diverse grid of hyperparameter combinations:

```{r }

xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  trees(),
  size = 10
)

xgb_grid
```

```{r}
ggplot(data = xgb_grid,
       aes(x = tree_depth, 
           y = min_n)) +
  geom_point(aes(color = factor(learn_rate), #coloring the bubbles based on learn_rate
                 size = trees), #size of the bubbles are based on the tress
             alpha = .5,
             show.legend = FALSE)
```

## 3. Model Tuning

```{r xgb_grid_result}

#install.packages("doParallel")
#install.packages("parallel")

library(doParallel)
library(parallel)

set.seed(76544)

#parallel processing
registerDoParallel(cores = parallel::detectCores()-1)

xgb_res <- tune_race_anova(object = xgb_spec,
                      preprocessor = weather_recipe,
                      resamples = resampling_foldcv,
                      grid = xgb_grid,
                      control = control_race(save_pred = TRUE))

stopImplicitCluster()

beepr::beep()

xgb_res$.metrics[[2]]

```

### 2025-04-26 code/Draft_script_No_Cat_Features.qmd
d

--- 2025-04-26 code/Draft_script_2.qmd
@@
 The following code chunk will keep the observations for site-years having latitude and longitude withing the Daymet range of co-ordinates.
+
+Declaration of AI use: the following code chunk was inspired and subsequently modified on the basis of code initially generated by ChatGPT
 
@@
 fe_month_sum <- fe_month %>%
-  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame
+  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "yield" in the group_by() to include it in the data frame because "yield" is our response variable so we must keep it in the data frame
   #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
@@
 #install.packages("xgboost") #new pacakage
+#install.packages("caret")
+
 library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
@@
 library(doParallel)   # For parallel computing (useful during resampling/tuning)
-#library(caret)       # Other great library for Machine Learning 
+library(caret)       # Other great library for Machine Learning 
 ```
@@
 # Create recipe for data preprocessing
-weather_recipe <- recipe(yield ~ ., data = weather_train) %>% # Remove identifier columns and months not in growing season
+weather_recipe <- recipe(yield ~ ., data = weather_train) %>% 
+  # Remove identifier columns and months not in growing season
   step_rm(
@@
     site,       # Remove site identifier
+    hybrid,     # Remove site identifier
     matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
@@
   # Turning every factor (prev_crop, planting_stages, harvest_stages, etc.) into dummies
-  step_dummy(all_nominal_predictors()) %>%
-  # Imputation for any numeric NAs   
-  step_impute_median(all_numeric_predictors())
+  step_dummy(all_nominal_predictors())
 
@@
   trees = tune(),
-  tree_depth = tune(),  # Maximum depth of each tree
-  min_n = tune(),  # Minimum samples required to split a node
-  learn_rate = tune()) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
+  tree_depth = tune() # ,  # Maximum depth of each tree
+  #min_n = tune(),  # Minimum samples required to split a node
+  #learn_rate = tune()
+  ) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
   set_engine("xgboost") %>% #specify engine 
   set_mode("regression")  # Set to mode
-   # Total number of boosting iterations
-
-        # Step size shrinkage for each boosting step
 
@@
 xgb_grid <- grid_latin_hypercube(
+  trees(),
   tree_depth(),
-  min_n(),
-  learn_rate(),
-  trees(),
+  #min_n(),
+  #learn_rate(),
   size = 10
@@
 ```
+
+Note: the following code chunk would not work if "tree_depth" and "learn_rate" were not fine tuned in the "xgb_spec" object.
 
@@
 #parallel processing
-registerDoParallel(cores = parallel::detectCores()-1)
+registerDoParallel(cores = parallel::detectCores()-1) #starts parallel processing
 
@@
 
-stopImplicitCluster()
+stopImplicitCluster() #ends parallel processing
 
@@
 ```
+
+## 4. Select Best Models
+
+We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:
+
+```{r}
+# Based on lowest RMSE
+best_rmse <- xgb_res %>% 
+  select_best(metric = "rmse")%>% 
+  mutate(source = "best_rmse")
+
+best_rmse
+
+```
+
+```{r}
+# Based on lowers RMSE within 1% loss
+best_rmse_pct_loss <- xgb_res %>% 
+  select_by_pct_loss("min_n",
+                     metric = "rmse",
+                     limit = 1
+                     )%>% 
+  mutate(source = "best_rmse_pct_loss")
+
+best_rmse_pct_loss
+```
+
+```{r}
+# Based on lowest RMSE within 1 se
+best_rmse_one_std_err <- xgb_res %>% 
+  select_by_one_std_err(metric = "rmse",
+                        eval_time = 100,
+                        trees
+                        )%>% 
+  mutate(source = "best_rmse_one_std_err")
+
+best_rmse_one_std_err
+```
+
+Here we use all three methods which we learn in this class for R2.
+
+```{r}
+# Based on greatest R2
+best_r2 <- xgb_res %>% 
+  select_best(metric = "rsq")%>% 
+  mutate(source = "best_r2")
+
+best_r2
+```
+
+```{r}
+# Based on lowers R2 within 1% loss
+best_r2_pct_loss <- xgb_res %>% 
+  select_by_pct_loss("min_n",
+                     metric = "rsq",
+                     limit = 1
+                     ) %>% 
+  mutate(source = "best_r2_pct_loss")
+
+best_r2_pct_loss
+```
+
+```{r}
+# Based on lowest R2 within 1 se
+best_r2_one_std_error <- xgb_res %>% 
+  select_by_one_std_err(metric = "rsq",
+                        eval_time = 100,
+                        trees
+                        ) %>%
+  mutate(source = "best_r2_one_std_error")
+
+best_r2_one_std_error
+```
+
+## Compare and Finalize Model
+
+```{r comparing values}
+best_rmse %>% 
+  bind_rows(best_rmse_pct_loss, 
+            best_rmse_one_std_err, 
+            best_r2, 
+            best_r2_pct_loss, 
+            best_r2_one_std_error)
+```
+
+## 5. Final Specification
+
+```{r final_spec_fit}
+final_spec <- boost_tree(
+  trees = best_r2$trees,           # Number of boosting rounds (trees)
+  tree_depth = best_r2$tree_depth #, # Maximum depth of each tree
+  #min_n = best_r2$min_n,           # Minimum number of samples to split a node
+  #learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
+) %>%
+  set_engine("xgboost") %>%
+  set_mode("regression")
+
+final_spec
+```
+
+## 6. Final Fit and Predictions
+
+## Validation
+
+```{r final_fit}
+set.seed(10)
+final_fit <- last_fit(final_spec,
+                weather_recipe,
+                split = weather_split)
+
+final_fit %>%
+  collect_predictions()
+```
+
+## 7. Evaluate on *Test Set*
+
+```{r final_fit_metrics}
+final_fit %>%
+  collect_metrics()
+```
+
+## 8. Evaluate on Training Set
+
+```{r}
+final_spec %>%
+  fit(yield ~ .,
+      data = bake(weather_prep, 
+                  weather_train)) %>%
+  augment(new_data = bake(weather_prep, 
+                          weather_train)) %>% 
+  rmse(yield, .pred) %>%
+  bind_rows(
+
+# R2
+final_spec %>%
+  fit(yield ~ .,
+      data = bake(weather_prep, 
+                  weather_train)) %>%
+  augment(new_data = bake(weather_prep, 
+                          weather_train)) %>% 
+  rsq(yield, .pred))
+```
+
+## 9. Predicted vs Observed Plot
+
+```{r}
+final_fit %>%
+  collect_predictions() %>%
+  ggplot(aes(x = yield,
+             y = .pred)) +
+  geom_point() +
+  geom_abline() +
+  geom_smooth(method = "lm") +
+  scale_x_continuous() +
+  scale_y_continuous() 
+```
+
+## 10. Variable Importance
+
+```{r final_spec}
+final_spec %>%
+  fit(yield ~ .,
+         data = bake(weather_prep, weather_train)) %>% #There little change in variable improtance if you use full dataset
+    vi() %>%
+  mutate(
+    Variable = fct_reorder(Variable, 
+                           Importance)
+  ) %>%
+  ggplot(aes(x = Importance, 
+             y = Variable)) +
+  geom_col() +
+  scale_x_continuous(expand = c(0, 0)) +
+  labs(y = NULL)
+```
+
+# End of XGboost
+
+# Start of Random Forest
+
+# Setup
+
+```{r}
+#| message: false
+#| warning: false
+
+#install.packages("ranger") #one of the packages in R that models random forest
+
+library(tidymodels)
+library(tidyverse) #takes a while to load
+library(vip)
+library(ranger)
+library(finetune)
+```
+
+## 2. Random Forest Training
+
+### a. Random Forest Model specification
+
+First, let's specify:  
+  - the **type of model** we want to train  
+  - which **engine** we want to use  
+  - which **mode** we want to use
+
+Random forest **hyperparameters**:  
+  - **trees**: number of trees in the forest    
+  - **mtry**: number of random features [or columns] sampled at each node split    
+    - **min_n**: minimum number of data points in a node that are required for the node to be split further
+
+[Important note: we are not gonna fine tune "min_n" in this exercise because of time constraint, but we should fine tune "min_n" to do random forest in our own data]
+
+Let's create a model specification that will **fine-tune** the first two for us.
+
+A given model type can be fit with different engines (e.g., through different packages). Here, we'll use the **ranger** engine/package.
+
+```{r rf_spec}
+
+rf_spec <- 
+  # Specifying random forest as our model type, asking to tune the hyperparameters
+  rand_forest(trees = tune(),
+              mtry = tune() #both "trees" and "mtry" are model-type level hyperparameters, meaning that we fine tune "trees" and "mtry" inside "rand_forest()"
+              ) %>%
+    # Specify the engine (= package)
+    set_engine("ranger") %>% #specifying "ranger" as the engine/package to run random forest 
+    # Specifying mode  
+    set_mode("regression") #random forest can handle both regression (when y is numerical) and classification (when y is categorical) #Here, we are specifying "set_mode("regression")" because our y variable is numerical [continuous]
+
+rf_spec
+
+```
+
+### b. Hyper-parameter tuning
+
+For our iterative search, we need:  
+  - Our model specification (`rf_spec`)  
+  - The recipe (`weather_recipe`)  
+  - Our **resampling strategy** (don't have yet)
+
+Let's define our resampling strategy below, using a 5-fold cross validation approach:
+
+```{r resampling_foldcv}
+
+set.seed(34549)
+
+rf_resampling_foldcv <- vfold_cv(weather_train, 
+                              v = 5) #but at least 10 folds are recommended
+
+rf_resampling_foldcv
+rf_resampling_foldcv$splits[[1]]
+rf_resampling_foldcv$splits[[2]]
+
+```
+
+We will use an iterative search algorithm called **simulated annealing**.
+
+In the algorithm below, we are asking for 10 iterations.
+
+```{r rf_grid_result}
+
+set.seed(76544)
+
+rf_grid_result <- tune_sim_anneal(object = rf_spec,
+                     preprocessor = weather_recipe,
+                     resamples = rf_resampling_foldcv,
+                    #param_info = rf_param,
+                    iter = 10 
+                     )
+
+beepr::beep()
+
+rf_grid_result
+rf_grid_result$.metrics[[2]]
+```
+
+Let's collect a summary of metrics (across all folds, for each iteration), and plot them.
+
+First, RMSE (lower is better):
+
+```{r RMSE}
+rf_grid_result %>%
+  collect_metrics() %>%
+  filter(.metric == "rmse") %>%
+  ggplot(aes(x = mtry, 
+             y = trees 
+             )) +
+  geom_path(group = 1) +
+  geom_point(aes(color = mean),
+             size = 3) + 
+  scale_color_viridis_b() +
+  geom_text(aes(label = .iter), nudge_x = .0005) +
+  labs(title = "RMSE")
+```
+
+Now, let's look into R2 (higher is better):
+
+```{r R2}
+
+rf_grid_result %>%
+  collect_metrics() %>%
+  filter(.metric == "rsq") %>%
+  ggplot(aes(x = mtry, 
+             y = trees 
+             )) +
+  geom_path(group = 1) +
+  geom_point(aes(color = mean),
+             size = 3) + 
+  scale_color_viridis_b() +
+  geom_text(aes(label = .iter), nudge_x = .0005) +
+  labs(title = "R2")
+
+```
+
+We want to see higher R2 (R2: higher is better).
+
+The following code chunk will select the best model based on "percent loss" of the metric.
+
+```{r}
+
+# Based on lowest RMSE
+best_rmse <- rf_grid_result %>%
+  select_by_pct_loss("trees",
+                     metric = "rmse",
+                     limit = 2 #"limit = 2": this is how many percent losses you are accepting, 2% in this case
+                     )
+
+best_rmse
+
+```
+
+The following code chunk will select the best model based on greatest "R2"
+
+```{r}
+
+# Based on greatest R2
+best_r2 <- rf_grid_result %>%
+  select_by_pct_loss("trees",
+                     metric = "rsq",
+                     limit = 2
+                     )
+
+best_r2
+
+```
+
+I got the he following results that were different than Dr. Bastos's results (because of the randomness of 2 hyperparameters going on at each tree for a lot of models):
+
+Based on RMSE, we would choose   
+  - mtry = 25 
+  - trees = 443
+
+Based on R2, we would choose   
+  - mtry = 22
+  - trees = 113
+
+Let's use the hyperparameter values that optimized R2 to fit our final model.
+
+```{r final_spec}
+final_spec <- rand_forest(trees = best_r2$trees,
+                          mtry = best_r2$mtry) %>%
+  # Specify the engine
+  set_engine("ranger",
+             importance = "permutation" #"permutation" is how we look into variable importance in random forest
+             ) %>%
+    # Specifying mode  
+  set_mode("regression")
+
+final_spec
+```
+
+## 3. Validation
+
+Now that we determined our best model, let's do our **last fit**.
+
+### What does last fit mean? (Will come in exam)
+
+This means 2 things:  
+  - Training the optimum hyperparameter values on the **entire training set**  
+  - Using it to **predict** on the **test set**
+
+These 2 steps can be completed in one function, as below:
+
+```{r final_fit}
+final_fit <- last_fit(final_spec, 
+                weather_recipe,
+                split = weather_split)
+
+final_fit %>%
+  collect_predictions()
+```
+
+Metrics on the **test set**:
+
+```{r}
+final_fit %>%
+  collect_metrics()
+```
+
+From the output,
+Fit metrics on the test set:
+rmse = 2.9317445	
+rsq	= 0.1882245
+
+Metrics on **train set** (for curiosity and compare to test set):  
+```{r}
+# RMSE
+final_spec %>%
+  fit(yield ~ .,
+      data = bake(weather_prep, 
+                  weather_train)) %>%
+  augment(new_data = bake(weather_prep, 
+                          weather_train)) %>% 
+  rmse(yield, .pred) %>%
+  bind_rows(
+
+    # R2
+    final_spec %>%
+      fit(yield ~ .,
+          data = bake(weather_prep, 
+                      weather_train)) %>%
+      augment(new_data = bake(weather_prep, 
+                              weather_train)) %>% 
+      rsq(yield, .pred)
+
+  )
+
+```
+
+From the output,
+Fit metrics on the train set:
+rmse = 1.1385078	
+rsq	= 0.9470296
+
+How does metrics on test compare to metrics on train?
+
+Fit metrics on the test set:
+rmse = 2.9317445	
+rsq	= 0.1882245
+
+Fit metrics on the train set:
+rmse = 1.1385078	
+rsq	= 0.9470296
+
+We see that the R2 is really r-e-a-l-l-y high in the train set (rsq	= 0.9470296), as compared to the R2 of the test set 0.1882245.
+The RMSE is lower in the train set (rmse = 1.1385078) as compared to the RMSE of the test set (rmse = 2.9317445).
+[This is the reason why it's okay to report the fit metrics of the train set to understand how well our model is explaining our data, but it is not a good measure of the predictive ability of the model.
+To understand the predictive ability of the model, we have to look at the fit metrics of the test set.
+Notice that the R2 of the test set was 0.1882245, whereas the R2 of the train set was 0.9470296 -- which is a very large difference.
+Hence, *we need to always report the fit metrics of the test set* along with the fit metrics of the train set.
+]
+
+Predicted vs. observed plot:  
+```{r}
+final_fit %>%
+  collect_predictions() %>%
+  ggplot(aes(x = yield,
+             y = .pred)) +
+  geom_point() +
+  geom_abline() +
+  geom_smooth(method = "lm") +
+  scale_x_continuous(limits = c(20, 40)) +
+  scale_y_continuous(limits = c(20, 40)) 
+```
+
+[**Very important note:**
+To interpret random forest model, we should always expcet to see (1) *perdicted vs.
+observed plot* and (3) *variable importance plot*, at the least -- these are the 2 things that must be mentioned in a random forest model.
+]
+
+*Variable importance:*
+
+[For Random Forest,] The importance metric we are evaluating here is **permutation**.
+
+In the permutation-based approach, for each tree, the out-of-bag sample is passed down the tree and the prediction accuracy is recorded. [as is]
+
+Then the values for each variable (one at a time) are randomly permuted and the accuracy is again computed.
+
+The decrease in accuracy as a result of this randomly shuffling of feature values is averaged over all the trees for each predictor.
+
+The variables with the **largest average decrease in accuracy** [after being permuted] are [the ones that are] considered **most important**.
+
+```{r}
+
+final_spec %>%
+  fit(yield ~ .,
+         data = bake(weather_prep, weather)) %>%
+    vi() %>%
+  mutate(
+    Variable = fct_reorder(Variable, 
+                           Importance)
+  ) %>%
+  ggplot(aes(x = Importance, 
+             y = Variable)) +
+  geom_col() +
+  scale_x_continuous(expand = c(0, 0)) +
+  labs(y = NULL)
+
+```
+
+#End of Random Forest
+
+#Start of getting yield prediction (yield_mg_ha) for the unseen data in "testing_submission.csv" file

### 2025-04-26 code/Draft_script_No_Cat_Features.qmd
d

--- 2025-04-26 code/Draft_script_2.qmd
@@
     grain_moisture = mean(grain_moisture, na.rm = TRUE),
-    date_planted   = first(date_planted),
-    date_harvested = first(date_harvested),
     .groups = "drop"
@@
     adjusted_yield = adjust_yield(yield_mg_ha, grain_moisture),
-    planting_date  = as.Date(date_planted,   "%m/%j/%y"),
-    harvest_date   = as.Date(date_harvested, "%m/%j/%y"),
-    plant_doy      = yday(planting_date),
-    harvest_doy    = yday(harvest_date),
   ) %>%
-  select(-yield_mg_ha, -grain_moisture, -date_planted, -date_harvested, -planting_date, -harvest_date) %>%
+  select(-yield_mg_ha, -grain_moisture) %>%
   ungroup()
@@
 
-meta_df2 <- meta_df %>%
+meta_clean <- meta_df %>%
   rename(
@@
   distinct(year, site, .keep_all = TRUE) %>%
+  select(-previous_crop) %>%
+  clean_names() %>%
   ungroup()
-
-meta_df2
-
-```
-
-The code below will create categories for previous crops.
-
-```{r}
-
-meta_clean <- meta_df2 %>%
-  mutate(
-    # 1) lowercase & trim
-    prev_crop_raw = str_to_lower(str_squish(previous_crop)),
-    # 2) recode into your desired levels
-    prev_crop = case_when(
-      prev_crop_raw == "cotton"                                   ~ "cotton",
-      prev_crop_raw == "peanut"                                   ~ "peanut",
-      prev_crop_raw %in% c("soybean", "soybeans")                 ~ "soybean",
-      prev_crop_raw == "sorghum"                                  ~ "sorghum",
-      prev_crop_raw %in% c("wheat", "winter wheat", "2019/20 wheat") ~ "wheat",
-      TRUE                                                         ~ "others"
-    ) %>%
-    # 3) make it a factor with the exact ordering you want
-    factor(levels = c("cotton", "peanut", "soybean", "sorghum", "wheat", "others"))
-  ) %>%
-  select(-previous_crop, -prev_crop_raw) %>%
-  clean_names()
 
@@
 ```
+
 ```{r}
@@
                 yield = adjusted_yield, 
-                prev_crop,
-                plant_doy,
-                harvest_doy,
                 soil.ph = soilpH, 
@@
 
-The following code chunk will feature engineer planting and harvest stages based on the following cut-offs:
-
-Planting
-•	early: DOY ≤ 120
-•	normal: 121–150
-•	late: > 150
-
-Harvest
-•	early: DOY ≤ 275
-•	normal: 276–305
-•	late: > 305
-
-```{r}
-
-fe_month <- fe_month%>%
-  mutate(
-    planting_stages = case_when(
-      plant_doy <= 120                     ~ "early",
-      plant_doy >= 121 & plant_doy <= 150  ~ "normal",
-      plant_doy > 150                      ~ "late"
-    ) %>% 
-      fct_relevel("early","normal","late"),
-
-    harvest_stages = case_when(
-      harvest_doy <= 275                    ~ "early",
-      harvest_doy >= 276 & harvest_doy <= 305 ~ "normal",
-      harvest_doy > 305                     ~ "late"
-    ) %>% 
-      fct_relevel("early","normal","late")
-  ) %>%
-  select(-plant_doy, -plant_doy)
-
-fe_month
-
-```
-
 The following code chunk will summarize daily weather variables based on month.
@@
 fe_month_sum <- fe_month %>%
-  group_by(year, site, hybrid, lon, lat, month_abb, yield, prev_crop, planting_stages, harvest_stages) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "strength_gtex" in the group_by() to include it in the data frame because "strength_gtex" is our response variable so we must keep it in the data frame
+  group_by(year, site, hybrid, month_abb, yield) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "yield" in the group_by() to include it in the data frame because "yield" is our response variable so we must keep it in the data frame
   #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
@@
   # Rounding to one decimal point
-  mutate(across(10:ncol(.), ~round(., 1) )) %>%
-  select(-lon, -lat, -mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)
+  mutate(across(c(4:ncol(.)), ~round(., 1) )) %>%
+  select( -mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)
 
 fe_month_sum_wide
-
-View(fe_month_sum_wide)
 
@@
 
-#write_csv(fe_month_sum_wide,
-          # "../data/weather_monthsum.csv")
+write_csv(fe_month_sum_wide,
+           "../data/weather_monthsum.csv")
 
@@
 library(doParallel)   # For parallel computing (useful during resampling/tuning)
-#library(caret)       # Other great library for Machine Learning 
+library(caret)       # Other great library for Machine Learning 
+```
+
+# Load the data set
+
+```{r weather}
+
+#weather <- read_csv("../data/weather_monthsum.csv")
+
+#weather
+
 ```
@@
 
-Creating a recipe is an easy way to port your processing steps for other data sets without needing to repeat code, and also only considering the data it is being applied to.
-
-You can find all available recipe step options here: https://tidymodels.github.io/recipes/reference/index.html
-
-Different model types require different processing steps.\
-Let's check what steps are required for an elastic net model (linear_reg). We can search for that in this link: https://www.tmwr.org/pre-proc-table #Don't need for numeric data
-
 ```{r weather_recipe}
@@
     site,       # Remove site identifier
+    hybrid,     # Remove site identifier
     matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
-  ) %>%
-  # Turning every factor (prev_crop, planting_stages, harvest_stages, etc.) into dummies
-  step_dummy(all_nominal_predictors()) %>%
-  # Imputation for any numeric NAs   
-  step_impute_median(all_numeric_predictors())
+  )
 
@@
 ```
+
 ## 2. Training
@@
   min_n = tune(),  # Minimum samples required to split a node
-  learn_rate = tune()) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
+  learn_rate = tune()
+  ) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
   set_engine("xgboost") %>% #specify engine 
   set_mode("regression")  # Set to mode
-   # Total number of boosting iterations
-
-        # Step size shrinkage for each boosting step
 
@@
 ```
+
 ### c. Hyperparameter grid with Latin Hypercube Sampling
@@
 xgb_grid <- grid_latin_hypercube(
+  trees(),
   tree_depth(),
@@
   learn_rate(),
-  trees(),
-  size = 10
+  size = 20
 )
@@
 xgb_grid
+
 ```
+
+Note: the following code chunk would not work if "tree_depth" and "learn_rate" were not fine tuned in the "xgb_spec" object.
 
@@
 #parallel processing
-registerDoParallel(cores = parallel::detectCores()-1)
+#registerDoParallel(cores = parallel::detectCores()-1) #starts parallel processing
 
@@
 
-stopImplicitCluster()
+#stopImplicitCluster() #ends parallel processing
 
@@
 ```
+
+## 4. Select Best Models
+
+We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:
+
+```{r}
+# Based on lowest RMSE
+best_rmse <- xgb_res %>% 
+  select_best(metric = "rmse")%>% 
+  mutate(source = "best_rmse")
+
+best_rmse
+
+```
+
+```{r}
+# Based on lowers RMSE within 1% loss
+best_rmse_pct_loss <- xgb_res %>% 
+  select_by_pct_loss("min_n",
+                     metric = "rmse",
+                     limit = 1
+                     )%>% 
+  mutate(source = "best_rmse_pct_loss")
+
+best_rmse_pct_loss
+```
+
+```{r}
+# Based on lowest RMSE within 1 se
+best_rmse_one_std_err <- xgb_res %>% 
+  select_by_one_std_err(metric = "rmse",
+                        eval_time = 100,
+                        trees
+                        )%>% 
+  mutate(source = "best_rmse_one_std_err")
+
+best_rmse_one_std_err
+```
+
+Here we use all three methods which we learn in this class for R2.
+
+```{r}
+# Based on greatest R2
+best_r2 <- xgb_res %>% 
+  select_best(metric = "rsq")%>% 
+  mutate(source = "best_r2")
+
+best_r2
+```
+
+```{r}
+# Based on lowers R2 within 1% loss
+best_r2_pct_loss <- xgb_res %>% 
+  select_by_pct_loss("min_n",
+                     metric = "rsq",
+                     limit = 1
+                     ) %>% 
+  mutate(source = "best_r2_pct_loss")
+
+best_r2_pct_loss
+```
+
+```{r}
+# Based on lowest R2 within 1 se
+best_r2_one_std_error <- xgb_res %>% 
+  select_by_one_std_err(metric = "rsq",
+                        eval_time = 100,
+                        trees
+                        ) %>%
+  mutate(source = "best_r2_one_std_error")
+
+best_r2_one_std_error
+```
+
+## Compare and Finalize Model
+
+```{r comparing values}
+best_rmse %>% 
+  bind_rows(best_rmse_pct_loss, 
+            best_rmse_one_std_err, 
+            best_r2, 
+            best_r2_pct_loss, 
+            best_r2_one_std_error)
+```
+
+## 5. Final Specification
+
+```{r final_spec_fit}
+final_spec <- boost_tree(
+  trees = best_r2$trees,           # Number of boosting rounds (trees)
+  tree_depth = best_r2$tree_depth, # Maximum depth of each tree
+  min_n = best_r2$min_n,           # Minimum number of samples to split a node
+  learn_rate = best_r2$learn_rate  # Learning rate (step size shrinkage)
+) %>%
+  set_engine("xgboost") %>%
+  set_mode("regression")
+
+final_spec
+```
+
+## 6. Final Fit and Predictions
+
+## Validation
+
+```{r final_fit}
+set.seed(10)
+final_fit <- last_fit(final_spec,
+                weather_recipe,
+                split = weather_split)
+
+final_fit %>%
+  collect_predictions()
+```
+
+## 7. Evaluate on *Test Set*
+
+```{r final_fit_metrics}
+final_fit %>%
+  collect_metrics()
+```
+
+## 8. Evaluate on Training Set
+
+```{r}
+final_spec %>%
+  fit(yield ~ .,
+      data = bake(weather_prep, 
+                  weather_train)) %>%
+  augment(new_data = bake(weather_prep, 
+                          weather_train)) %>% 
+  rmse(yield, .pred) %>%
+  bind_rows(
+
+# R2
+final_spec %>%
+  fit(yield ~ .,
+      data = bake(weather_prep, 
+                  weather_train)) %>%
+  augment(new_data = bake(weather_prep, 
+                          weather_train)) %>% 
+  rsq(yield, .pred))
+```
+
+## 9. Predicted vs Observed Plot
+
+```{r}
+final_fit %>%
+  collect_predictions() %>%
+  ggplot(aes(x = yield,
+             y = .pred)) +
+  geom_point() +
+  geom_abline() +
+  geom_smooth(method = "lm") +
+  scale_x_continuous() +
+  scale_y_continuous() 
+```
+
+## 10. Variable Importance
+
+```{r final_spec}
+final_spec %>%
+  fit(yield ~ .,
+         data = bake(weather_prep, weather_train)) %>% #There little change in variable improtance if you use full dataset
+    vi() %>%
+  mutate(
+    Variable = fct_reorder(Variable, 
+                           Importance)
+  ) %>%
+  ggplot(aes(x = Importance, 
+             y = Variable)) +
+  geom_col() +
+  scale_x_continuous(expand = c(0, 0)) +
+  labs(y = NULL)
+```
+
+# End of XGboost

### 2025-04-26 code/draft_random_forest.qmd
d

---
title: "Data Science Applied to Ag - Final Project - ML"
format:
  html:
    embed-resources: true
    toc: true
    theme: cerulean
author: Md Shakir Moazzem, Umar Munir
---

# Start of data wrangling

# Introduction

This script contains the data wrangling steps for the final project.

# Setup

##Loading packages

The following code chunk will load necessary packages.

```{r Setup, message=F}

# Installing packages

#install.packages("tidyverse")
#install.packages("readxl")
#install.packages("janitor")
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("readr")
#install.packages("lubridata")
#install.packages("stringr")

# Loading packages

library(tidyverse)
library(readxl) # to read excel files
library(janitor) # to clean data; helps fix and standardize the column names
library(dplyr) # wrangling
library(tidyr) # wrangling
library(readr) # to export csv
library(lubridate)
library(stringr)

```

# Setup  
```{r}
#| message: false
#| warning: false

#install.packages("ranger") #one of the packages in R that models random forest

library(tidymodels)
library(tidyverse) #takes a while to load
library(vip)
library(ranger)
library(finetune)
```

```{r weather}
weather <- read_csv("../data/weather_monthsum.csv")

weather
```

# ML workflow  
We're going to use the same workflow as we used for elastic net.

## 1. Pre-processing  
Here's where we perform **data split** and **data processing**.

### a. Data split  
For data split, let's use **70% training / 30% testing**.

```{r weather_split}
# Setting seed to get reproducible results  
set.seed(931735)

# Setting split level  
weather_split <- initial_split(weather, 
                               prop = .7,
                               strata = strength_gtex) # strata = strength_gtex: to do stratified sampling based on strength_gtex to make sure the distribution of strength across training and testing are similar

weather_split
```

```{r weather_train}
# Setting train set 
weather_train <- training(weather_split)

weather_train
```

```{r weather_test}
# Setting test split
weather_test <- testing(weather_split)

weather_test
```

```{r distribution}
ggplot() +
  geom_density(data = weather_train, 
               aes(x = strength_gtex),
               color = "red") +
  geom_density(data = weather_test, 
               aes(x = strength_gtex),
               color = "blue")

```

```{r weather_recipe}
weather_recipe <-
  # Defining predicted and predictor variables
  recipe(strength_gtex ~ .,
         data = weather_train) %>%
  # Removing year and site  
    step_rm(year, site, matches("Jan|Feb|Mar|Apr|Nov|Dec")) #%>%
  # Normalizing all numeric variables except predicted variable
  #step_normalize(all_numeric(), -all_outcomes())

weather_recipe
```

```{r weather_prep}
weather_prep <- weather_recipe %>%
  prep()

weather_prep
```

A given model type can be fit with different engines (e.g., through different packages). Here, we'll use the **ranger** engine/package.

```{r rf_spec}

rf_spec <- 
  # Specifying random forest as our model type, asking to tune the hyperparameters
  rand_forest(trees = tune(),
              mtry = tune() #both "trees" and "mtry" are model-type level hyperparameters, meaning that we fine tune "trees" and "mtry" inside "rand_forest()"
              ) %>%
    # Specify the engine (= package)
    set_engine("ranger") %>% #specifying "ranger" as the engine/package to run random forest 
    # Specifying mode  
    set_mode("regression") #random forest can handle both regression (when y is numerical) and classification (when y is categorical) #Here, we are specifying "set_mode("regression")" because our y variable is numerical [continuous]

rf_spec

```

### b. Hyper-parameter tuning  
For our iterative search, we need:  
  - Our model specification (`rf_spec`)  
  - The recipe (`weather_recipe`)  
  - Our **resampling strategy** (don't have yet)

> Notice that for "random forest", we do not need to specify the parameter information, as we needed for CIT. The reason is that for random forest, all hyperparameters to be tuned are specified at the model level, whereas for CIT: one hyperparameter was at model level ["tree_depth" was fine-tuned inside decision_tree() as:  decision_tree(tree_depth = tune()) ] and another hyperparameter was at the engine level ["conditional_min_criterion" was fine tuned inside set_engine() as: set_engine("partykit", conditional_min_criterion = tune())]. Engine level hyperparameters need to be "finalized" and have their range of values set up before they can be used in search methods.

[Note: CIT had one hyperparameter at the level of the model, and another hyperparameter at the level of the engine. Because of that, we had to do 1 extra step for the hyperparameter at the level of the engine to set the levels that would be searched in the grid. In random forest, both of the hyperparameters are fine tuned at the level of the model i.e., inside "rand_forest( trees = tune(), mtry = tune() )", and we don't have any hyperparameter at the engine level. Therefore, we can skip one of the steps here in random forest that were required for CIT.]

> We used 10-fold CV before. It took about 10-min to run the grid on my side, so to avoid a long wait time in class, let's switch to 5-fold CV this time around. But we always need to do at least 10 folds of our data.

Let's define our resampling strategy below, using a 5-fold cross validation approach:  
```{r resampling_foldcv}
set.seed(34549)
resampling_foldcv <- vfold_cv(weather_train, 
                              v = 5) #but at least 10 folds are recommended

resampling_foldcv
resampling_foldcv$splits[[1]]
resampling_foldcv$splits[[2]]
```

In the algorithm below, we are asking for 50 iterations.

```{r rf_grid_result}
set.seed(76544)
rf_grid_result <- tune_sim_anneal(object = rf_spec,
                     preprocessor = weather_recipe,
                     resamples = resampling_foldcv,
                    #param_info = rf_param,
                    iter = 10 #we are using "iter = 10" because of time constraint in class #but at least 100 is recommended #"iter = 10" also means that the iteration is gonna run 10 times to find the best #If we specified "iter = 100", the iteration would run 100 times to find the best, running 100 iteration is the best practice
                     )

beepr::beep() #plays a bell sound when the code chunk finishes running #this comes in handy for a code chunk that takes a long time to run, because the bell sound indicates the code chunk has finished running

rf_grid_result
rf_grid_result$.metrics[[2]]
```

First, RMSE (lower is better):
```{r RMSE}
rf_grid_result %>%
  collect_metrics() %>%
  filter(.metric == "rmse") %>%
  ggplot(aes(x = mtry, 
             y = trees 
             )) +
  geom_path(group = 1) +
  geom_point(aes(color = mean),
             size = 3) + 
  scale_color_viridis_b() +
  geom_text(aes(label = .iter), nudge_x = .0005) +
  labs(title = "RMSE")
```

Now, let's look into R2 (higher is better):

```{r R2}
rf_grid_result %>%
  collect_metrics() %>%
  filter(.metric == "rsq") %>%
  ggplot(aes(x = mtry, 
             y = trees 
             )) +
  geom_path(group = 1) +
  geom_point(aes(color = mean),
             size = 3) + 
  scale_color_viridis_b() +
  geom_text(aes(label = .iter), nudge_x = .0005) +
  labs(title = "R2")

```

> Previously, we selected the single best model. Now, let's select the best model by "percent loss" of the metric, so we choose a model among the top ones that is more parsimonious.

```{r}
# Based on lowest RMSE
best_rmse <- rf_grid_result %>%
  select_by_pct_loss("trees",
                     metric = "rmse",
                     limit = 2 #"limit = 2": this is how many percent losses you are accepting, 2% in this case
                     )

best_rmse

```

```{r}
# Based on greatest R2
best_r2 <- rf_grid_result %>%
  select_by_pct_loss("trees",
                     metric = "rsq",
                     limit = 2
                     )

best_r2

```

# 3. Validation

Now that we determined our best model, let's do our **last fit**.

### What does last fit mean? (Will come in exam)

This means 2 things:  
  - Training the optimum hyperparameter values on the **entire training set**  
  - Using it to **predict** on the **test set**

These 2 steps can be completed in one function, as below:

```{r final_fit}
final_fit <- last_fit(final_spec, 
                weather_recipe,
                split = weather_split)

final_fit %>%
  collect_predictions()
```

Metrics on the **test set**:

```{r}
final_fit %>%
  collect_metrics()
```

Metrics on **train set** (for curiosity and compare to test set):  
```{r}
# RMSE
final_spec %>%
  fit(strength_gtex ~ .,
      data = bake(weather_prep, 
                  weather_train)) %>%
  augment(new_data = bake(weather_prep, 
                          weather_train)) %>% 
  rmse(strength_gtex, .pred) %>%
  bind_rows(

    # R2
    final_spec %>%
      fit(strength_gtex ~ .,
          data = bake(weather_prep, 
                      weather_train)) %>%
      augment(new_data = bake(weather_prep, 
                              weather_train)) %>% 
      rsq(strength_gtex, .pred)

  )

```

Predicted vs. observed plot:  
```{r}
final_fit %>%
  collect_predictions() %>%
  ggplot(aes(x = strength_gtex,
             y = .pred)) +
  geom_point() +
  geom_abline() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(20, 40)) +
  scale_y_continuous(limits = c(20, 40)) 
```

The variables with the **largest average decrease in accuracy** [after being permuted] are [the ones that are] considered **most important**.

```{r}

final_spec %>%
  fit(strength_gtex ~ .,
         data = bake(weather_prep, weather)) %>%
    vi() %>%
  mutate(
    Variable = fct_reorder(Variable, 
                           Importance)
  ) %>%
  ggplot(aes(x = Importance, 
             y = Variable)) +
  geom_col() +
  scale_x_continuous(expand = c(0, 0)) +
  labs(y = NULL)

```

### 2025-04-26 code/02_xgboost.qmd
d

--- 2025-04-26 code/Draft_script_No_Cat_Features.qmd
@@
 ---
-title: "Data Science Applied to Ag - Final Project - ML"
-format:
-  html:
-    embed-resources: true
-    toc: true
-    theme: cerulean
-author: Md Shakir Moazzem, Umar Munir
+title: "02_xgboost"
+format: html
+editor: visual
 ---
-
-# Start of data wrangling
 
@@
 
-This script contains the data wrangling steps for the final project.
+This script contains ML Workflow with XGBoost
 
-# Setup
+## Loading libraries
 
-##Loading packages
-
-The following code chunk will load necessary packages.
-
-```{r Setup, message=F}
-
-# Installing packages
-
-#install.packages("tidyverse")
-#install.packages("readxl")
-#install.packages("janitor")
-#install.packages("dplyr")
-#install.packages("tidyr")
-#install.packages("readr")
-#install.packages("lubridata")
-#install.packages("stringr")
-
-# Loading packages
-
-library(tidyverse)
-library(readxl) # to read excel files
-library(janitor) # to clean data; helps fix and standardize the column names
-library(dplyr) # wrangling
-library(tidyr) # wrangling
-library(readr) # to export csv
-library(lubridate)
-library(stringr)
-
-```
-
-## Reading data
-
-The following code chunk will read the csv files for the 3 training data sets
-
-```{r training data import, message=F, warning=F}
-
-#reading the csv files for the 3 training data sets
-
-trait_df <- read_csv("../data/training/training_trait.csv") 
-meta_df  <- read_csv("../data/training/training_meta.csv")
-soil_df  <- read_csv("../data/training/training_soil.csv")
-
-```
-
-## EDA
-
-```{r}
-
-summary(trait_df)
-
-View(trait_df)
-
-```
-```{r}
-
-summary(meta_df)
-
-```
-
-```{r}
-
-summary(soil_df)
-
-```
-
-# Data wrangling on "trait_df"
-
-The code below creates a function to adjust the yield_mg_ha for 15.5% grain moisture.
-
-```{r}
-
-# Function to transform yield to 15.5% moisture
-adjust_yield <- function(yield_mg_ha, grain_moisture) {
-  yield_mg_ha * (100 - grain_moisture) / (100 - 15.5)
-}
-
-```
-
-The code below conducts data wrangling on "trait_df"
-
-```{r clean_and_summarize_trait, message=F, warning=F}
-
-trait_clean <- trait_df %>%
-  select(-block) %>%
-  mutate(
-    site = str_remove_all(site, "[a-z]"), # removing all lowercase letters from site
-    site = str_replace(site, "-.*$", ""), # keeping only text before any dash
-    site = str_replace(site, "_.*$", "") # keeping only text before any underscore
-  ) %>%
-  group_by(year, site, hybrid) %>%
-  summarize(
-    yield_mg_ha    = mean(yield_mg_ha, na.rm = TRUE),
-    grain_moisture = mean(grain_moisture, na.rm = TRUE),
-    .groups = "drop"
-  ) %>%
-  mutate(
-    adjusted_yield = adjust_yield(yield_mg_ha, grain_moisture),
-  ) %>%
-  select(-yield_mg_ha, -grain_moisture) %>%
-  ungroup()
-
-trait_clean
-
-```
-
-```{r exporting wrangled data set}
-
-#write_csv(trait_avg_df,
-          #"../data/training_trait_clean.csv")
-```
-
-# Data wrangling on "soil_df"
-
-The following code chunk will conduct data wrangling on  "training_soil.csv"
-
-```{r data_wrangling_training_soil, message=F, warning=F}
-
-soil_clean <- soil_df %>%
-  mutate(
-    # 1) remove all lowercase letters
-    site = str_remove_all(site, "[a-z]"),
-    # 2) keep only text before any dash
-    site = str_replace(site, "-.*$", ""),
-    # 3) keep only text before any underscore
-    site = str_replace(site, "_.*$", "")
-  ) %>%
-  ungroup()
-
-soil_clean
-```
-
-```{r exporting wrangled data set}
-
-#write_csv(soil_clean,
-          #"../data/training_soil_clean.csv")
-```
-
-# Data wrangling on "meta_df"
-
-The following code chunk will conduct data wrangling on  "training_meta.csv".
-
-```{r clean_meta_sites}
-
-meta_clean <- meta_df %>%
-  rename(
-    lon = longitude,
-    lat = latitude
-  ) %>%
-  mutate(
-    # 1) remove all lowercase letters
-    site = str_remove_all(site, "[a-z]"),
-    # 2) keep only text before any dash
-    site = str_replace(site, "-.*$", ""),
-    # 3) keep only text before any underscore
-    site = str_replace(site, "_.*$", "")
-  ) %>%
-  distinct(year, site, .keep_all = TRUE) %>%
-  select(-previous_crop) %>%
-  clean_names() %>%
-  ungroup()
-
-meta_clean
-
-```
-
-# EDA for cleaned data
-
-```{r}
-summary(trait_clean)
-```
-
-```{r}
-summary(soil_clean)
-```
-
-```{r}
-summary(meta_clean)
-```
-
-```{r exporting wrangled data set}
-
-#write_csv(meta_clean,
-          #"../data/training_meta_clean.csv")
-```
-
-# Merging all 3 cleaned data frames
-
-```{r}
-
-merged_clean  <- trait_clean %>%
-  left_join(soil_clean, by = c("year", "site")) %>%
-  left_join(meta_clean, by = c("year", "site"))
-
-merged_clean
-```
-
-```{r}
-summary(merged_clean)
-```
-
-```{r exporting merged data set}
-
-#write_csv(merged_clean,
-          #"../data/training_merged_clean.csv")
-```
-
-# End of Data Wrangling
-
-#Start of Open Source Daymet weather data download
-
-The following code chunk will load necessary packages.
-
-```{r Setup, message=F, warning=F}
-
-# Installing packages
-
-#install.packages("tidyverse")
-#install.packages("sf") #to manipulate vector geospatial files
-#install.packages("daymetr") #to retrieve data from daymet database website through R
-#install.packages("remotes") #to install R packages that are not available on CRAN, and available on GitHub
-#remotes::install_github("ropensci/USAboundaries") 
-#remotes::install_github("ropensci/USAboundariesData")
-
-# Loading packages
-
-library(tidyverse) #need to load "tidyverse" package at first
-library(sf) # for US map #to manipulate vector geo-spatial points
-library(daymetr) #to retrieve data from daymet database website through R
-library(remotes)
-library(USAboundaries) # for US state boundaries
-library(USAboundariesData)
-
-```
-
-The following code chunk will create a map of the USA and plot the sites in the map based on their latitude and longitude.
-
-```{r create map of USA and add points, message=F, warning=F}
-
-states <- us_states() %>% 
-  filter( !(state_abbr %in% c("PR", "AK", "HI")) ) #to remove "PR" (Puerto Rico), "AK" (Alaska), and "HI" (Hawaii)
-
-ggplot() +
-  geom_sf(data = states) + #"geom_sf()" is used to plot "sf" object, which we just created above as "states" object; plots all states and territories of USA
-  geom_point(data = merged_clean,
-             aes(x = lon, #"Longitude" goes on longitude
-                 y = lat) #"Latitude" goes on latitude
-             ) +
-  labs(
-    title = "Corn Trial Site Locations (2014–2023)",
-    x     = "Longitude",
-    y     = "Latitude"
-  )
-
-```
-
-The following code chunk will keep the observations for site-years having latitude and longitude withing the Daymet range of co-ordinates.
-
-Declaration of AI use: the following code chunk was inspired and subsequently modified on the basis of code initially generated by ChatGPT
-
-```{r}
-
-# Define Daymet bounding box (WGS-84)
-min_lat <- 14.53
-max_lat <- 52.00
-min_lon <- -131.104
-max_lon <- -52.95
-
-# Filter merged to Daymet’s valid range, dropping any NA coords
-merged_daymet <- merged_clean %>%
-  filter(
-    !is.na(lat),
-    !is.na(lon),
-    lat  >= min_lat,
-    lat  <= max_lat,
-    lon  >= min_lon,
-    lon  <= max_lon
-  )
-
-# Report how many rows remain (and were dropped)
-message("Rows kept: ", nrow(merged_daymet), 
-        " (dropped: ", nrow(merged_clean) - nrow(merged_daymet), ")")
-
-```
-The following code chunk will extract unique combinations of year, site, and their coordinates.
-
-```{r unique_site_years_with_coords}
-
-site_year_df <- merged_daymet %>%
-  select(year, site, lon, lat) %>%  # need to include longitude and latitude along with site-years
-  distinct() %>%                    
-  arrange(year, site)
-
-site_year_df
-
-```
-
-The following code chunk will download the weather data for all unique combinations of year, site, and their coordinates in the "site_year_df" object.
-
-```{r}
-
-weather_daymet_all <- site_year_df %>% 
-  mutate(weather = pmap(list(.y = year, 
-                             .site = site, 
-                             .lat = lat, 
-                             .lon = lon), 
-                        function(.y, .site, .lat, .lon) 
-                          download_daymet( 
-                            site = .site, 
-                            lat = .lat, #specifying ".lat" placeholder for "lat = " argument
-                            lon = .lon, #specifying ".lon" placeholder for "lon = " argument
-                            start = .y, 
-                            end = .y, 
-                            simplify = T,
-                            silent = T) %>% #end of " download_daymet()" function
-                          rename(.year = year,
-                                 .site = site) 
-                        ))
-
-weather_daymet_all
-
-```
-
-The following code chunk will unnest the weather column.
-
-```{r}
-
-weather_daymet_unnest <- weather_daymet_all %>%
-  select(year, site, weather) %>% 
-  unnest(weather) %>% 
-  pivot_wider(names_from = measurement, 
-              values_from = value) %>% 
-  janitor::clean_names()
-
-weather_daymet_unnest
-
-View(weather_daymet_unnest)
-
-```
-
-# Merging the weather data retrieved from Daymet to the "merged_clean" data
-
-The following code chunk will merge the weather data retrieved from Daymet to the "merged_clean" data
-
-```{r merge_weather_with_data}
-
-# Join daily weather onto your cleaned trial data by year & site
-
-daymet_all_unnest <- merged_clean %>%
-  left_join(
-    weather_daymet_unnest,
-    by = c("year", "site")
-  )
-
-fieldweather <- daymet_all_unnest
-
-fieldweather
-
-View(fieldweather)
-
-```
-
-## Exporting the merged data
-
-```{r}
-
-#write_csv(daymet_all_unnest,
-          #"../data/merged_fieldweatherdata.csv"
-          #)
-
-```
-
-# End of retreiving weather data from Daymet
-
-# Start of feature engineering
-
-## Setup
-
-```{r}
-
-#install.packages("ggridges")
-
-library(ggridges)
-library(tidyverse)
-
-```
-
-```{r}
-
-#fieldweather <- read_csv("../data/merged_fieldweatherdata.csv")
-
-#fieldweather
-
-```
-
-The following code chunk will keep desired variables that we will use further and get abbreviated month name based on the date.
-
-```{r}
-
-fe_month <- fieldweather %>%
-  # Selecting needed variables
-  dplyr::select(year, 
-                site, 
-                hybrid, 
-                lon, 
-                lat, 
-                yday, 
-                yield = adjusted_yield, 
-                soil.ph = soilpH, 
-                soil.om.pct = om_pct, 
-                soil.k.ppm = soilk_ppm, 
-                soil.p.ppm = soilp_ppm, 
-                dayl.s = dayl_s, #to rename variable name from "dayl_s" to dayl.s
-                prcp.mm = prcp_mm_day, #to rename variable name to "prcp.mm"
-                srad.wm2 = srad_w_m_2,#to rename variable name to "srad.wm2"
-                tmax.c = tmax_deg_c, #to rename variable name to "tmax.c"
-                tmin.c = tmin_deg_c,#to rename variable name to "tmin.c"
-                vp.pa = vp_pa #to rename variable name to "vp.pa"
-                ) %>%
-  # Creating a date class variable  
-  mutate(date_chr = paste0(year, "/", yday)) %>% 
-  mutate(date = as.Date(date_chr, "%Y/%j")) %>% 
-  # Extracting month from date  
-  mutate(month = month(date)) %>% 
-  mutate(month_abb = month(date, label = T)) #To get abbreviated month name e.g., Jan, Feb, Mar,...,Dec #month_abb is in "ord" (ordinal) format
-
-fe_month
-```
-
-The following code chunk will summarize daily weather variables based on month.
-
-```{r fe_month_sum}
-
-fe_month_sum <- fe_month %>%
-  group_by(year, site, hybrid, month_abb, yield) %>% #If we do a summarise() after group_by(), any column that's not in the group_by() is gone, so we need to include "yield" in the group_by() to include it in the data frame because "yield" is our response variable so we must keep it in the data frame
-  #Because we are gonna be applying a "summarize()" function to different columns, we are gonna use a function called "across()" 
-  summarise(across(.cols = c(soil.ph, 
-                             soil.om.pct, 
-                             soil.k.ppm, 
-                             soil.p.ppm,
-                             dayl.s,
-                             srad.wm2,
-                             tmax.c,
-                             tmin.c,
-                             vp.pa),
-                   .fns = mean, #do not indicate the actual function "mean()", just use the word "mean"
-                   .names = "mean_{.col}"), #specifying the weather variables that we want their mean as new column variables #1st across() is applying the "mean" function (to summarize "mean")
-            across(.cols = prcp.mm,
-                   .fns = sum,
-                   .names = "sum_{.col}"
-                   ) #specifying the weather variable (prcp.mm) that we want its sum as new column variable #2nd across() is summarizing sum #2nd across() is applying the "sum" function (to summarize "sum")
-            ) %>%
-  ungroup() #To convert from "group" to "tibble"
-
-fe_month_sum
-
-```
-
-The following code chunk will check "tmax.c" and "prcp.mm" for the first site-year and month for the purpose of double checking to make sure that we did everything okay the way we intended.
-
-```{r}
-
-fe_month %>%
-  filter(year == 2014 & 
-           site == "DEH1" &
-           hybrid == "B37/MO17" &
-           month_abb == "Jan") %>%
-  summarise(tmax.c = mean(tmax.c),
-            prcp.mm = sum(prcp.mm))
-
-```
-
-The code below will put the month as part of the column name.
-
-```{r fe_month_sum_wide}
-
-fe_month_sum_wide <- fe_month_sum %>%
-  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% 
-  mutate(varname = paste0(name, "_", month_abb)) %>% 
-  dplyr::select(-name, -month_abb) %>% 
-  pivot_wider(names_from = varname,
-              values_from = value) %>%
-  # Rounding to one decimal point
-  mutate(across(c(4:ncol(.)), ~round(., 1) )) %>%
-  select( -mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)
-
-fe_month_sum_wide
-
-```
-
-# EDA round 2  
-Let's make a ridge plot to visualize the distribution of one variable over months.
-
-```{r, message=FALSE, warning=FALSE}
-
-#install.packages("ggridges")
-library(ggridges) #really powerful package to plot distributions e.g., density plot
-
-ggplot(data = fe_month_sum,
-       aes(x = mean_tmax.c,
-           y = month_abb,
-           fill = stat(x) #fill = stat(x): specific to "ggridges"
-           )
-       ) +
-  geom_density_ridges_gradient(scale = 3,
-                               rel_min_height = 0.01) + #"the 2nd argument is"rel_min_height = 0.01": to cut down the tails of the distribution to look nice
-  scale_fill_viridis_c(option = "C") + #"viridis"scale_fill_viridis_c()": color-blind-friendly color scale #argument (option = "C") to change the color to "magma" because magma is weather related; "option = " varies from A to F
-  theme(legend.position = "none") #"theme(legend.position = "none")": to remove the legend from the graph
-
-```
-
-Now let's do that for all weather variables [ we will automate using map2() to conduct iteration ]
-
-```{r}
-
-finalplots <- fe_month_sum %>%
-  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% #we need to do pivot_longer() to iterate over the weather variables
-  group_by(name) %>%
-  nest() %>% #for iteration, we use "group_by()" followed by "nest()" in combo
-  #we will use map2() since we need to iterate over 2 columns
-  mutate(plot = map2(data, name, #map2() takes 2 arguments: the 1st argument becomes .x, the 2nd argument becomes .y #we must use map2() with a mutate() at first for iteration #"map2(data, name" : we want to iterate over "name" for "data"
-                     ~ ggplot( data = .x, # .x represent "data" in the map2() function, so we need to use data = .x [= "data" from map2() ] to feed the "data" from map2() function as the data of ggplot() #.x is the iterating column of map2() that is a place holder for the 1st argument of map2()
-       aes(x = value,
-           y = month_abb,
-           fill = stat(x)
-           )
-       ) +
-  geom_density_ridges_gradient(scale = 3,
-                               rel_min_height = 0.01) + 
-  scale_fill_viridis_c(option = "C") + 
-  theme(legend.position = "none") +
-  labs(x = .y) #to rename the x-axis as the variable name # .y is the placeholder for "name" (i.e., variable name) in map2() function
-                     ))
-
-finalplots
-
-```
-
-```{r}
-
-finalplots$plot #$plot to print all the ggplots for each variables
-
-```
-
-```{r}
-
-write_csv(fe_month_sum_wide,
-           "../data/weather_monthsum.csv")
-
-```
-
-# End of feature engineering
-
-# Start of XGboost
-
-# ML Workflow with XGBoost
-
-## Step 1: Load Libraries
-
-We begin by loading the required libraries:
+The following code chunk loads the required libraries.
 
@@
 #| warning: false
+
+#install.packages("tidymodels")   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
+#install.packages("finetune")     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
+#install.packages("vip")          # For plotting variable importance from fitted models
+#install.packages("xgboost")      # XGBoost implementation in R
+#install.packages("ranger")       # Fast implementation of Random Forests
+#install.packages("tidyverse")    # Data wrangling and visualization
+#install.packages("doParallel")   # For parallel computing (useful during resampling/tuning)
+#install.packages("caret")  
 #install.packages("xgboost") #new pacakage
@@
 
-# Load the data set
+## Loading the data set
+
+The following code chunk will load the "weather_monthsum.csv" data set
 
@@
 
-#weather <- read_csv("../data/weather_monthsum.csv")
+weather <- read_csv("../data/weather_monthsum.csv")
 
-#weather
-
-```
-
-The following code chunk will rename the "fe_month_sum_wide" data frame as "weather".
-
-```{r}
-
-weather <- fe_month_sum_wide
 weather
@@
 ## 1. Pre-processing
+
+The following code chunk will conduct pre-processing.
 
@@
 
-For data split, let's use **70% training / 30% testing**.
+The following code chunks will conduct data split (70% training / 30% testing).
 
@@
 
-### b. Distribution of target variable
+### b. Distribution of target variable "yield"
+
+The following code chunk will create a density plot of target variable "yield"
 
 ```{r distribution}
+
 ggplot() +
@@
 ### c. Data processing with recipe
+
+The following code chunk will conduct data processing with recipe
 
@@
 
+The following code chunk will prep the recipe to estimate any required statistics.
+
 ```{r weather_prep}
@@
 
-First, let's specify:\
-- the **type of model** we want to train\
-- which **engine** we want to use\
-- which **mode** we want to use
-
-> Elastic nets can only be run for a numerical response variable.
-XgBoost can be run with either numerical (regression) or categorical (classification) explanatory variable.
-Therefore, we have the need to specify the mode here.
-
-XgBoost **hyperparameters**:
-
--   **`trees`**: The number of boosting rounds (i.e., how many trees will be added sequentially).\
--   **`tree_depth`**: Controls how deep each individual tree can grow. Deeper trees can capture more complex interactions but also increase the risk of overfitting.\
--   **`min_n`**: Minimum number of observations required in a node for it to be split. Acts as a regularization tool to prevent overly specific splits.\
--   **`learn_rate`**: Also known as `eta`, it controls how much each additional tree contributes to the overall model. Smaller values make the model more stable but require more trees.
+The following code chunk will fine tune the XgBoost hyperparameters.
 
@@
 
-We use 5-fold cross-validation to evaluate model performance during tuning:
+The following code chunk will conduct 5-fold cross-validation to evaluate model performance during tuning.
 
@@
 
-We use Latin hypercube sampling to generate a diverse grid of hyperparameter combinations:
+The following code chunk will use Latin hypercube sampling to generate a diverse grid of hyperparameter combinations.
 
@@
 
-Note: the following code chunk would not work if "tree_depth" and "learn_rate" were not fine tuned in the "xgb_spec" object.
+The following code chunk will plot the hyperparameter combinations.
 
@@
 ## 3. Model Tuning
+
+The following code chunk will conduct model tuning.
 
@@
 
-We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:
+We select the best models using three strategies (lowest RMSE, highest R2, within 1 SE, within 2% loss).
+
+The following code chunk will select best model based on lowest RMSE.
 
 ```{r}
+
 # Based on lowest RMSE
@@
 
+The following code chunk will select best model based on lowest RMSE within 1% loss
+
 ```{r}
-# Based on lowers RMSE within 1% loss
+# Based on lowest RMSE within 1% loss
 best_rmse_pct_loss <- xgb_res %>% 
@@
 ```
+
+The following code chunk will select best model based on lowest RMSE within 1 SE.
 
@@
 
-Here we use all three methods which we learn in this class for R2.
+The following code chunk will select best model based on greatest R2.
 
@@
 
+The following code chunk will select best model based on greatest R2 within 1% loss.
+
 ```{r}
-# Based on lowers R2 within 1% loss
+# Based on greatest R2 within 1% loss
 best_r2_pct_loss <- xgb_res %>% 
@@
 
+The following code chunk will select best model based on greatest R2 within 1 SE
+
 ```{r}
-# Based on lowest R2 within 1 se
+# Based on greatest R2 within 1 se
 best_r2_one_std_error <- xgb_res %>% 
@@
 
+The following code chunk will compare all models
+
 ```{r comparing values}
@@
 ## 5. Final Specification
+
+The following code chunk will conduct final specification.
 
@@
 
+The following code chunk will conduct validation.
+
 ```{r final_fit}
@@
 
+The following code chunk will evaluate fit metrics on the test set.
+
 ```{r final_fit_metrics}
@@
 ## 8. Evaluate on Training Set
+
+The following code chunk will evaluate fit metrics on the training set.
 
@@
 
+The following code chunk will create a Predicted vs Observed Plot.
+
 ```{r}
@@
 ## 10. Variable Importance
+
+The following code chunk will create a Variable Importance plot.
 
@@
 ```
-
-# End of XGboost

### 2025-04-26 code/Draft_script_No_Cat_Features.qmd
d

@@
 
+The following code chunk will conduct model tuning.
+
 ```{r xgb_grid_result}
@@
 
-We select the best models using three strategies (best, within 1 SE, within 2% loss) which we learned in class:
-
-```{r}
+We select the best models using three strategies (lowest RMSE, highest R2, within 1 SE, within 2% loss).
+
+The following code chunk will select best model based on lowest RMSE.
+
+```{r}
+
 # Based on lowest RMSE
@@
 
-```{r}
-# Based on lowers RMSE within 1% loss
+The following code chunk will select best model based on lowest RMSE within 1% loss
+
+```{r}
+# Based on lowest RMSE within 1% loss
 best_rmse_pct_loss <- xgb_res %>% 
@@
 
-```{r}
-# Based on lowers R2 within 1% loss
+The following code chunk will select best model based on greatest R2 within 1% loss.
+
+```{r}
+# Based on greatest R2 within 1% loss
 best_r2_pct_loss <- xgb_res %>% 
@@
 
-```{r}
-# Based on lowest R2 within 1 se
+The following code chunk will select best model based on greatest R2 within 1 SE
+
+```{r}
+# Based on greatest R2 within 1 se
 best_r2_one_std_error <- xgb_res %>% 
@@
 ## Compare and Finalize Model
+
+The following code chunk will compare all models
 
@@
 
+The following code chunk will conduct final specification.
+
 ```{r final_spec_fit}
@@
 ## 9. Predicted vs Observed Plot
+
+The following code chunk will create a Predicted vs Observed Plot.
 
@@
 
+The following code chunk will create a Variable Importance plot.
+
 ```{r final_spec}
@@
 
-# End of XGboost
+# Obtaining "yield" prediction on the "testing_submission.csv" data set
+
+```{r}
+
+# Loading libraries 
+library(tidyverse)    # dplyr, purrr, tidyr, readr, stringr, etc.
+library(lubridate)    # dates → yday, month
+library(janitor)      # clean_names()
+library(daymetr)      # download_daymet()
+library(tidymodels)   # recipe(), bake(), fit(), predict()
+# (assumes final_spec and weather_prep already exist in your environment)
+
+# Reading & cleaning testing data
+
+submission   <- read_csv("../data/testing/testing_submission.csv") %>% 
+  mutate(site = str_remove_all(site, "[a-z]"),
+         site = str_replace(site, "-.*$", ""),
+         site = str_replace(site, "_.*$", ""))
+
+meta_test    <- read_csv("../data/testing/testing_meta.csv")  %>% 
+  rename(lon = longitude, lat = latitude) %>% 
+  mutate(site = str_remove_all(site, "[a-z]"),
+         site = str_replace(site, "-.*$", ""),
+         site = str_replace(site, "_.*$", "")) %>% 
+  distinct(year, site, .keep_all = TRUE) %>% 
+  select(-previous_crop)
+
+soil_test    <- read_csv("../data/testing/testing_soil.csv")  %>% 
+  mutate(site = str_remove_all(site, "[a-z]"),
+         site = str_replace(site, "-.*$", ""),
+         site = str_replace(site, "_.*$", ""))
+
+test_base <- submission %>%
+  left_join(soil_test, by = c("year","site")) %>%
+  left_join(meta_test, by = c("year","site"))
+
+# Downloading Daymet weather for 2024 site‐years 
+site_year_test <- test_base %>% 
+  select(year, site, lon, lat) %>% 
+  distinct()
+
+weather_test <- site_year_test %>%
+  mutate(weather = pmap(
+    list(.y = year, .site = site, .lat = lat, .lon = lon),
+    ~ download_daymet(site = .site,
+                      lat   = .lat,
+                      lon   = .lon,
+                      start = .y,
+                      end   = .y,
+                      simplify = TRUE,
+                      silent = TRUE)
+  )) %>%
+  select(year, site, weather) %>%
+  unnest(weather) %>%
+  pivot_wider(names_from = measurement, values_from = value) %>%
+  clean_names()
+
+test_full <- test_base %>%
+  left_join(weather_test, by = c("year","site"))
+
+# Feature engineering monthly summaries for 2024 site‐years
+fe_test <- test_full %>%
+  # create date & month
+  mutate(date = as.Date(paste0(year, "/", yday), "%Y/%j"),
+         month_abb = month(date, label = TRUE)) %>%
+  # select and rename to match training
+  transmute(
+    year, site, hybrid,
+    month_abb,
+    soil.ph       = soilp_h,    # adjust these names if your clean_names() differ
+    soil.om.pct   = om_pct,
+    soil.k.ppm    = soilk_ppm,
+    soil.p.ppm    = soilp_ppm,
+    dayl.s        = dayl_s,
+    prcp.mm       = prcp_mm_day,
+    srad.wm2      = srad_w_m_2,
+    tmax.c        = tmax_deg_c,
+    tmin.c        = tmin_deg_c,
+    vp.pa         = vp_pa
+  ) %>%
+  group_by(year, site, hybrid, month_abb) %>%
+  summarise(
+    mean_soil.ph     = mean(soil.ph,   na.rm = TRUE),
+    mean_soil.om.pct = mean(soil.om.pct, na.rm = TRUE),
+    mean_soil.k.ppm  = mean(soil.k.ppm,  na.rm = TRUE),
+    mean_soil.p.ppm  = mean(soil.p.ppm,  na.rm = TRUE),
+    mean_dayl.s      = mean(dayl.s,      na.rm = TRUE),
+    sum_prcp.mm      = sum(prcp.mm,      na.rm = TRUE),
+    mean_srad.wm2    = mean(srad.wm2,    na.rm = TRUE),
+    mean_tmax.c      = mean(tmax.c,      na.rm = TRUE),
+    mean_tmin.c      = mean(tmin.c,      na.rm = TRUE),
+    mean_vp.pa       = mean(vp.pa,       na.omit = TRUE)
+  ) %>%
+  ungroup() %>%
+  pivot_longer(
+    cols = starts_with(c("mean_", "sum_")),
+    names_to = "name",
+    values_to = "value"
+  ) %>%
+  mutate(varname = paste0(name, "_", month_abb)) %>%
+  select(-name, -month_abb) %>%
+  pivot_wider(names_from = varname, values_from = value) %>%
+  select(-ends_with("_NA"))
+
+# Prepping test features & prediction 
+test_prepped  <- bake(weather_prep, new_data = fe_test)
+
+# fitting final model on all training data
+final_model   <- final_spec %>% 
+  fit(yield ~ ., data = bake(weather_prep, weather))
+
+# generating predictions
+predictions   <- predict(final_model, new_data = test_prepped)
+
+# binding back 
+submission_pred <- fe_test %>% 
+  select(year, site, hybrid) %>% 
+  bind_cols(predictions)
+
+write_csv(submission_pred,
+          "../data/testing/testing_submission_pred.csv")
+
+```

### 2025-04-26 code/Draft_script_No_Cat_Features.qmd
d

@@
 
-# Loading libraries 
-library(tidyverse)    # dplyr, purrr, tidyr, readr, stringr, etc.
-library(lubridate)    # dates → yday, month
-library(janitor)      # clean_names()
-library(daymetr)      # download_daymet()
-library(tidymodels)   # recipe(), bake(), fit(), predict()
-# (assumes final_spec and weather_prep already exist in your environment)
-
 # Reading & cleaning testing data

### 2025-04-26 code/Draft_script_No_Cat_Features.qmd
d

@@
   pivot_wider(names_from = measurement, values_from = value) %>%
-  clean_names()
+  #clean_names()
 
@@
 fe_test <- test_full %>%
-  # create date & month
+  # creating date & month
   mutate(date = as.Date(paste0(year, "/", yday), "%Y/%j"),
          month_abb = month(date, label = TRUE)) %>%
-  # select and rename to match training
+  # selecting and renaming to match training
   transmute(
@@
     month_abb,
-    soil.ph       = soilp_h,    # adjust these names if your clean_names() differ
+    soil.ph       = soilp_h,    # will have to adjust these names if the clean_names() differ
     soil.om.pct   = om_pct,
